dyngen: a multi-modal simulator for spearheading new single-cell omics analyses

### 2021-05-12 manuscript/manuscript.Rmd
update

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- Robrecht Cannoodt*
- Wouter Saelens*
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, results="hold")
library(tidyverse)
```

@@

# Main text
<!-- format: 1500 Words, no headings, max. 2 figures, max. 20 references-->
Single-cell simulation engines are becoming increasingly important for testing and benchmarking computational methods, a pressing need in the widely expanding field of single-cell biology. Complementary to real biological data, synthetic data provides a valuable alternative where the actual ground truth is completely known and thus can be compared to, in order to make quantitative evaluations of computational methods that aim to reconstruct this ground truth [@zappia_splattersimulationsinglecell_2017]. In addition, simulation engines are more flexible when it comes to stress-testing computational methods, for example by varying the parameters of the simulation, such as the amount of noise, samples, and cells measured, allowing benchmarking of methods over a wide range of possible scenarios. In this way, they can even guide the design of real biological experiments, finding out the best conditions to be used as input for subsequent computational pipelines.

@@

Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. In addition, dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and single-cell knockdown experiments). For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Table\ \ref{tab:comparison}).

@@

<!-- RESULTS -->
We demonstrate dyngen's broad applicability by evaluating three novel types of computational approaches for which no simulation engines exist yet: cell-specific network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}). We emphasize that our main aim here is to illustrate the potential of dyngen for these evaluations, rather than performing large-scale benchmarking, which would require assessing many more quantitative and qualitative aspects of each method [@weber_essentialguidelinescomputational_2019].

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
    \caption{
      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).}
      \textbf{A:} Trajectory alignment aligns two trajectories between samples. We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP).
      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving. We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell. We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR). We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
    }
    \label{fig:applications}
\end{figure}

**Trajectory alignment** methods align trajectories from different samples and allow studying the differences between the different trajectories. For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient. Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of a trajectory could be aligned as well. Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018]. cellAlign [@alpert_alignmentsinglecelltrajectories_2018] uses DTW to perform trajectory alignment, but also includes interpolation and scaling of the single cell data as a preprocessing step. We evaluate the performance of DTW and cellAlign by simulating 40 datasets, each containing two linear trajectories generated with the same gene regulatory network but with slightly different simulation kinetics. We assess the accuracy of the obtained alignments by comparing the generated alignment path with the worst possible alignment that could be performed (ABWAP). A visual guide for this metric can be found in Figure \ \ref{fig:traj_align}D. Overall, cellAlign performs significantly better than DTW (Figure\ \ref{fig:traj_align}, which is likely due to the interpolation and scaling steps provided by cellAlign, reducing noise in the data and improving the comparability of the trajectories. Note that, in this comparison, only linear trajectory alignment is performed. While dyngen can generate non-linear trajectories (e.g. cyclic or branching), both aligning non-linear trajectories and constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.

**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018]. Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2020]. Yet, to date, no quantitative assessment of their accuracy has been performed, mainly due to the difficulty in obtaining real ground-truth data to do so. In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen simulation, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded at any particular point in time. We executed velocyto and scvelo (with 2 different parameter settings, stochastic and dynamical) on 42 datasets with a variety of backbones (including linear, bifurcating, cyclic, disconnected). We evaluated the predictions using two metrics (Figure\ \ref{fig:velocity}), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine"). While both velocyto and scvelo obtained high scores for the velocity arrow cosine metric (overall 25th percentile = 0.606), the velocity correlation is rather low (overall 75th percentile = 0.156). This means that predicting the RNA velocity (i.e. transcription rate minus the decay rate) for any particular gene is challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory. In terms of velocity correlation, no method performed significantly better than the other, whereas "scvelo stochastic" performed slightly worse than "scvelo dynamical" and velocyto in terms of velocity arrow cosine score.

**Cell-specific network inference** (CSNI) methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state. While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of their performance is until now lacking. This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods. Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present. We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017]. For each dataset, we computed the mean AUROC and AUPR scores of the individual cells. Comparing the mean AUROC and AUPR showed that pySCENIC significantly outperforms both LIONESS and SSN, and in turn that LIONESS significantly outperforms SSN. The poor performance of SSN is expected, as its methodology for predicting a cell-specific is simply computing the difference in Pearson correlation values applied to the whole dataset and the whole dataset minus one sample. This strategy performs poorly in large datasets where cell correlations are high, as the removal of one cell will not yield large differences in correlation values and will result in mostly noise. Overall, pySCENIC almost always performs better than LIONESS, except for a few datasets where LIONESS does manage to obtain a higher AUROC score. However, by using a different internal network inference (e.g. GENIE3 [@huynh-thu_inferringregulatorynetworks_2010] or pySCENIC's GRNBoost2 [@moerman_grnboost2arboretoefficient_2019]) could significantly increase the performance obtained by LIONESS.

<!-- DISCUSSION -->
In summary, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference. In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015]. Additionally, one could anticipate technological developments in single-cell multi-omics. In this way, dyngen allows designing and evaluating the performance and robustness of new types of computational analyses before experimental data becomes available, comparing which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis. One major assumption of dyngen is that cells are assumed to be well-mixed and independent from each other. Subdividing a cell into multiple 2D or 3D subvolumes or allowing cells to exchange molecules, respectively, could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].

# Data Availability
Source data for box plots in Figures\ \ref{fig:applications}, \ref{fig:traj_align}, \ref{fig:velocity} and \ref{fig:scgrn} are provided with this paper. All code and data required to reproduce the analysis are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript). The datasets generated for the different use cases are available on Zenodo with record number 4637926 (doi: [10.5281/zenodo.4637926](https://doi.org/10.5281/zenodo.4637926)).

# Code Availability
dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen). The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).

@@

# Methods {#sec:dyngen-methods}
The workflow to generate _in silico_ single-cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).

@@

Several examples of module networks are given in Figure\ \ref{fig:example_backbones_onlymodules}.
A simple chain of modules (where one module upregulates the next) results in a _linear_ process. By having the last module repress the first module, the process becomes _cyclic_. Two modules repressing each other is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process. Finally, a _converging_ process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.

@@

## Generating the gene regulatory network {#sec:dyngen-grn}
The GRN is generated based on the given module network in four main steps (Figure\ \ref{fig:gen_feature_network}).

**Step 1, sampling the transcription factors (TF).** The TFs are the main drivers of the molecular changes in the simulation. The user provides a backbone and the number of TFs to generate. Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$). A TF inherits the 'burn' and 'basal expression' from the module it belongs to.

**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone. These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone. A TF can only be regulated by other TFs or itself.

**Step 3, sampling the target subnetwork.**
A user-defined number of target genes are added to the GRN. Target genes are regulated by a TF or another target gene, but are always downstream of at least one TF. To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently existing TFs are mapped to regulators in the FANTOM5 GRN. The targets are drawn from the FANTOM5 GRN weighted by their page rank value, to create an induced GRN. For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$). The interactions connecting a target gene and its regulators are added to the GRN.

**Step 4, sampling the housekeeping subnetwork.**
Housekeeping genes are completely separate from any TFs or target genes. A user-defined set of housekeeping genes is also sampled from the FANTOM5 GRN. The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$). A random gene is sampled and a breadth-first-search is performed to sample the desired number of housekeeping genes.

@@

The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle). Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\x G$. Splicing turns one pre-mRNA $\x G$ into a mature mRNA $\x G$. Translation uses a mature mRNA $\y G$ to produce a protein $\z G$. Pre-mRNA, mRNA and protein degradation results in the removal of a $\x G$, $\y G$, and $\z G$ molecule, respectively.

The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a per-gene constant (Table\ \ref{tab:reaction_params}). The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs. The per-gene and per-interaction constants are based on the median reported production-rates and half-lives of molecules measured of 5000 mammalian genes [@schwanhausser_globalquantificationmammalian_2011], except that the transcription rate has been amplified by a factor of 10.

@@

\begin{align}
  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
\end{align}

@@

Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.

GillespieSSA2 is an optimised library for performing SSA simulations. The propensity functions are compiled to C++ and SSA approximations can be used which allow triggering many reactions simultaneously at each iteration. The framework also allows storing the abundance levels of molecules only after a specific interval has passed since the previous census. By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information. In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.

@@

**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Figure\ \ref{fig:sample_cells}A). From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.

**Time series** Assuming that the final time of the simulation is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$. $N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$. By default, $k = 8$ and $g = 0.75$. For usual dyngen simulations, $10 \leq T \leq 20$. For larger values of $T$, $k$ and $g$ should be increased accordingly.

@@

## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure\ \ref{fig:example_backbones}). Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another. For each edge, a simulation is run using the end state of an upstream branch as the initial expression vector, and only allowing the modules as predefined by the attribute to change.

As an example, consider the cyclic trajectory shown in Figure\ \ref{fig:example_backbones}. State S0 begins with an expression vector of all zero values. To simulate the transition from S0 to S1, regulation of the genes in modules A, B and C are turned on. After a predefined period of time, the end state of this transition is considered the expression vector of state S1. To simulate the transition from S1 to S2, regulation of the genes in modules D and E are turned on, while the regulation of genes in module C is turned off. During this simulation, the expression of genes in modules A, B, D, and E is thus allowed to change. The end state of the simulation is considered the expression vector of state S2.

>>>>

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- name: Robrecht Cannoodt
  institute: [dambi, twist, di]
  equal_contributor: true
- name: Wouter Saelens
  institute: [dambi, twist, epfl]
  equal_contributor: true
- name: Louise Deconinck
  institute: [dambi, twist]
- name: Yvan Saeys
  institute: [dambi, twist]
  correspondence: true
  email: yvan.saeys@ugent.be
institute: 
  - dambi: Data Mining and Modelling for Biomedicine group, VIB Center for Inflammation Research, Ghent, Belgium
  - twist: Department of Applied Mathematics, Computer Science, and Statistics, Ghent University, Ghent, Belgium
  - di: Data Intuitive, Lebbeke, Belgium
  - epfl: Institute of Bioengineering, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
    pandoc_args:
      - '--lua-filter=lua/scholarly-metadata.lua'
      - '--lua-filter=lua/author-info-blocks.lua'
biblatexoptions: sorting=none,url=false,backend=bibtex
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

```{r setup, include=FALSE}
 # for lua files regarding author information, see https://stackoverflow.com/a/52919256
knitr::opts_chunk$set(collapse=TRUE, results="hold")
library(tidyverse)
```

@@

# Introduction
<!-- format: 1500 Words, no headings, max. 2 figures, max. 20 references-->
Single-cell simulation engines are becoming increasingly important for testing and benchmarking computational methods, a pressing need in the widely expanding field of single-cell biology. Complementary to real biological data, synthetic data provides a valuable alternative where the actual ground truth is completely known and thus can be compared to, in order to make quantitative evaluations of computational methods that aim to reconstruct this ground truth [@zappia_splattersimulationsinglecell_2017]. In addition, simulation engines are more flexible when it comes to stress-testing computational methods, for example by varying the parameters of the simulation, such as the amount of noise, samples, and cells measured, allowing benchmarking of methods over a wide range of possible scenarios. In this way, they can even guide the design of real biological experiments, finding out the best conditions to be used as input for subsequent computational pipelines.

@@

Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. In addition, dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and single-cell knockdown experiments). For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Supplementary Table\ 1).

@@

# Results
We demonstrate dyngen's broad applicability by evaluating three novel types of computational approaches for which no simulation engines exist yet: cell-specific network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}). We emphasize that our main aim here is to illustrate the potential of dyngen for these evaluations, rather than performing large-scale benchmarking, which would require assessing many more quantitative and qualitative aspects of each method [@weber_essentialguidelinescomputational_2019].

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
    \caption{
      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).} Box plots denote the $Q_0$ to $Q_4$ quartile values.
      \textbf{A:} Trajectory alignment aligns two trajectories between samples. We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP). 
      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving. We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell. We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR). We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
    }
    \label{fig:applications}
\end{figure}

Use-case "trajectory alignment". Trajectory alignment methods align trajectories from different samples and allow studying the differences between the different trajectories. For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient. Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of a trajectory could be aligned as well. Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018]. cellAlign [@alpert_alignmentsinglecelltrajectories_2018] uses DTW to perform trajectory alignment, but also includes interpolation and scaling of the single cell data as a preprocessing step. We evaluate the performance of DTW and cellAlign by simulating 40 datasets, each containing two linear trajectories generated with the same gene regulatory network but with slightly different simulation kinetics. We assess the accuracy of the obtained alignments by comparing the generated alignment path with the worst possible alignment that could be performed (Supplementary Fig. \ 1D), named the Area Between Worst And Prediction (ABWAP) score. Overall, cellAlign performs significantly better than DTW (Supplementary Fig.\ 1), which is likely due to the interpolation and scaling steps provided by cellAlign, reducing noise in the data and improving the comparability of the trajectories. Note that, in this comparison, only linear trajectory alignment is performed. While dyngen can generate non-linear trajectories (e.g. cyclic or branching), both aligning non-linear trajectories and constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.

Use-case "RNA velocity". RNA velocity methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018]. Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2020]. Yet, to date, no quantitative assessment of their accuracy has been performed, mainly due to the difficulty in obtaining real ground-truth data to do so. In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen simulation, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded at any particular point in time. We executed velocyto and scvelo (with 2 different parameter settings, stochastic and dynamical) on 42 datasets with a variety of backbones (including linear, bifurcating, cyclic, disconnected). We evaluated the predictions using two metrics (Supplementary Fig.\ 2), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine"). While both velocyto and scvelo obtained high scores for the velocity arrow cosine metric (overall 25th percentile = 0.606), the velocity correlation is rather low (overall 75th percentile = 0.156). This means that predicting the RNA velocity (i.e. transcription rate minus the decay rate) for particular individual genes can be challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory. In terms of velocity correlation, no method performed significantly better than the other, whereas "scvelo stochastic" performed slightly worse than "scvelo dynamical" and velocyto in terms of velocity arrow cosine score. Note that, given that some genes are more informative in determining the overall directionality of cell progression, performing a feature selection before computing the embedded dimensionality reduction might result in significantly improved velocity arrow cosine scores.

Use-case "Cell-specific network inference" (CSNI). CSNI methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state. While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of their performance is until now lacking. This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods. Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present. We used this ground-truth to compare the performance of three CSNI methods (Supplementary Fig.\ 3): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017]. For each dataset, we computed the mean AUROC and AUPR scores of the individual cells. Comparing the mean AUROC and AUPR showed that pySCENIC significantly outperforms both LIONESS and SSN, and in turn that LIONESS significantly outperforms SSN. The poor performance of SSN is expected, as its methodology for predicting a cell-specific is simply computing the difference in Pearson correlation values applied to the whole dataset and the whole dataset minus one sample. This strategy performs poorly in large datasets where cell correlations are high, as the removal of one cell will not yield large differences in correlation values and will result in mostly noise. Overall, pySCENIC almost always performs better than LIONESS, except for a few datasets where LIONESS does manage to obtain a higher AUROC score. However, by using a different internal network inference (e.g. GENIE3 [@huynh-thu_inferringregulatorynetworks_2010] or pySCENIC's GRNBoost2 [@moerman_grnboost2arboretoefficient_2019]) could significantly increase the performance obtained by LIONESS.

# Discussion
dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference. In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015]. Additionally, one could anticipate technological developments in single-cell multi-omics. In this way, dyngen allows designing and evaluating the performance and robustness of new types of computational analyses before experimental data becomes available, comparing which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis. One major assumption of dyngen is that cells are assumed to be well-mixed and independent from each other. Subdividing a cell into multiple 2D or 3D subvolumes or allowing cells to exchange molecules, respectively, could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].

@@

# Methods {#sec:dyngen-methods}
The workflow to generate _in silico_ single-cell data consists of six main steps (Supplementary Fig\ 4).

@@

Several examples of module networks are given in Supplementary Fig.\ 5.
A simple chain of modules (where one module upregulates the next) results in a _linear_ process. By having the last module repress the first module, the process becomes _cyclic_. Two modules repressing each other is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process. Finally, a _converging_ process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.

@@

## Generating the gene regulatory network {#sec:dyngen-grn}
The GRN is generated based on the given module network in four main steps (Supplementary Fig.\ 6).

Step 1, sampling the transcription factors (TF). The TFs are the main drivers of the molecular changes in the simulation. The user provides a backbone and the number of TFs to generate. Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$). A TF inherits the 'burn' and 'basal expression' from the module it belongs to.

Step 2, generating the TF interactions. Let each TF be regulated according to the interactions in the backbone. These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone. A TF can only be regulated by other TFs or itself.

Step 3, sampling the target subnetwork. A user-defined number of target genes are added to the GRN. Target genes are regulated by a TF or another target gene, but are always downstream of at least one TF. To sample the interactions between target genes, one of the many FANTOM5 [@lizio_gatewaysfantom5promoter_2015] GRNs is sampled. The currently existing TFs are mapped to regulators in the FANTOM5 GRN. The targets are drawn from the FANTOM5 GRN weighted by their page rank value, to create an induced GRN. For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$). The interactions connecting a target gene and its regulators are added to the GRN.

Step 4, sampling the housekeeping subnetwork. Housekeeping genes are completely separate from any TFs or target genes. A user-defined set of housekeeping genes is also sampled from the FANTOM5 GRN. The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$). A random gene is sampled and a breadth-first-search is performed to sample the desired number of housekeeping genes.

@@

The effects of each reaction mimic the respective biological processes (Supplementary Table\ 2, middle). Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\x G$. Splicing turns one pre-mRNA $\x G$ into a mature mRNA $\x G$. Translation uses a mature mRNA $\y G$ to produce a protein $\z G$. Pre-mRNA, mRNA and protein degradation results in the removal of a $\x G$, $\y G$, and $\z G$ molecule, respectively.

The propensity of all reactions except transcription are all linear functions (Supplementary Table\ 2, right) of the abundance level of some molecule multiplied by a per-gene constant (Supplementary Table\ 3). The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs. The per-gene and per-interaction constants are based on the median reported production-rates and half-lives of molecules measured of 5000 mammalian genes [@schwanhausser_globalquantificationmammalian_2011], except that the transcription rate has been amplified by a factor of 10.

@@

\begin{align}
  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} 
\end{align}

@@

Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j \textrm{prop}_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $\textrm{prop}_j$ the propensity value of the $j$th reaction for the current state of the simulation.

GillespieSSA2 is an optimised library for performing SSA simulations. The propensity functions are compiled to C++ and SSA approximations can be used which allow triggering many reactions simultaneously at each iteration. The framework also allows storing the abundance levels of molecules only after a specific interval has passed since the previous census. By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information. In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained.

@@

**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Supplementary Fig.\ 7A). From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.

**Time series** Assuming that the final time of the simulation is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$. $N_i = N / k$ cells are sampled uniformly from each interval (Supplementary Fig.\ 7B), rounded such that $\sum N_i = N$. By default, $k = 8$ and $g = 0.75$. For usual dyngen simulations, $10 \leq T \leq 20$. For larger values of $T$, $k$ and $g$ should be increased accordingly.

@@

## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Supplementary Fig.\ 8). Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another. For each edge, a simulation is run using the end state of an upstream branch as the initial expression vector, and only allowing the modules as predefined by the attribute to change.

As an example, consider the cyclic trajectory shown in Supplementary Fig.\ 8. State S0 begins with an expression vector of all zero values. To simulate the transition from S0 to S1, regulation of the genes in modules A, B and C are turned on. After a predefined period of time, the end state of this transition is considered the expression vector of state S1. To simulate the transition from S1 to S2, regulation of the genes in modules D and E are turned on, while the regulation of genes in module C is turned off. During this simulation, the expression of genes in modules A, B, D, and E is thus allowed to change. The end state of the simulation is considered the expression vector of state S2.
